---
title: "Assimilate production under water scarcity, EC, pH and salinity build-up"
subtitle: "A structured literature review of crop- and field-level ODE models"
author: "Hans Schepers"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  eval: false
  echo: true
bibliography: null
---

::: {.callout-important}
## Provenance of this review

This review was written **from memory (model training), without an internet search**, per
standing instruction. Titles, journals and years are reported as recalled; DOI links are
given so each can be verified in one click. Treat the reference list as a *checklist to
verify*, not as a validated bibliography. Where I am unsure of a detail I say so explicitly
in the text.
:::

# Scope and the common state-space frame

## What is being reviewed

Models that compute **assimilate production at the canopy / field / greenhouse-compartment
level** — i.e. gross or net carbon gain of a whole stand — and that make that production
depend on:

1. **water scarcity** (matric limitation, hydraulic limitation, stomatal closure),
2. **EC / osmotic potential** of the root-zone solution,
3. **salinity build-up** in soil or substrate (a slow state, not a driver),
4. **pH / alkalinity** of soil or nutrient solution,
5. **soil or substrate type** (retention curve, buffering, sodicity response).

Explicitly *out of scope*: biochemical detail below the leaf (Calvin cycle intermediates,
mesophyll conductance partitioning, Rubisco activation kinetics). The Farquhar–von
Caemmerer–Berry (FvCB) leaf model appears only as a plug-in submodel where a reviewed
model uses it.

## The state-space frame used throughout

Every model below is re-expressed in the frame

$$
\frac{dX}{dt} = f\big(X,\; p,\; u(t),\; w(t)\big), \qquad
Y(t) = g\big(X,\; p,\; u(t),\; w(t)\big),
$$

with the control generated *inside* the same system by a rule-based policy that has its own
perception states:

$$
u(t) = h_r\big(X,\; p_u,\; w(t)\big).
$$

| Symbol | Role | Typical members in this literature | Typical units |
|---|---|---|---|
| $X$ | states | crop dry mass, leaf area index, root-zone water, root-zone salt, leaf sodium, substrate pH / alkalinity, sensor perception states | g DM m^-2^; m^2^ m^-2^; mm; g m^-2^; mmol~c~ m^-2^; pH unit; – |
| $Y$ | outputs | harvestable yield, drainage volume, drain EC, transpiration, water productivity | g DM m^-2^; mm d^-1^; dS m^-1^; mm d^-1^; g DM L^-1^ |
| $p$ | parameters | radiation use efficiency, extinction coefficient, specific leaf area, salinity 50 % point, buffer capacity | g DM MJ^-1^; –; m^2^ g^-1^; dS m^-1^; mmol~c~ m^-2^ pH-unit^-1^ |
| $u$ | controls | irrigation rate, target leaching fraction, supply EC set point, acid dose, pH set point, CO~2~ set point, screen position | mm d^-1^; –; dS m^-1^; mmol~c~ m^-2^ d^-1^; –; umol mol^-1^; – |
| $w$ | uncontrolled drivers | global radiation, reference evapotranspiration, air temperature, vapour pressure deficit, rainfall, irrigation-water EC and alkalinity | MJ m^-2^ d^-1^; mm d^-1^; degC; kPa; mm d^-1^; dS m^-1^; mmol~c~ L^-1^ |

Note on the prefix convention used here: `u` is used for *micro* (umol, uS cm^-1^) and the
symbol $u(t)$ is used for control. Context disambiguates; in code, control variables always
carry the word `control_` or an explicit name such as `irrigation_rate`.

## Structural remark: the canopy is a von Bertalanffy system

Worth stating up front, because it organises the whole review. Canopy carbon balance has
exactly the *supply-surface minus maintenance-mass* structure:

$$
\frac{dW}{dt} \;=\; a\!\Big(q(\cdot),\; \text{assimilation capacity}\Big) \;-\; \text{maintenance} \cdot W .
$$

- The **gain term is a surface term**: light capture saturates with leaf area,
  $1 - \exp(-k \cdot \mathrm{LAI})$, which is the canopy analogue of $W^{2/3}$ — a
  projected-area scaling, not a mass scaling.
- The **loss term is a mass term**: maintenance respiration is proportional to standing
  biomass (and to a $Q_{10}$-type temperature factor).
- The gate $a(\cdot)$ is again a **smoothed minimum** of supply and demand: source-limited
  (light, water, CO~2~) versus sink-limited (temperature-driven expansion, fruit load).
- The food-quality function $q(\cdot)$ maps onto **solution quality**: EC, ion ratios, pH,
  and the frequency/duration of irrigation events (the "meal pattern" of a drip-irrigated
  crop is literally a meal-frequency problem).

Where crop models differ from animal von Bertalanffy models is the *water* channel: water
scarcity acts on the surface term through stomatal closure **and** shrinks the surface
itself (leaf expansion, senescence) with different time constants. That split is the single
most important structural difference among the models below.

# How assimilate production is formulated: three backbones

The whole literature reduces to three ways of writing the gain term. They are not
equivalent, and they respond to water and salt stress differently.

## Backbone A — radiation-driven (RUE / LUE)

$$
\frac{d\,\text{dry mass}}{dt} =
\text{radiation use efficiency} \cdot \text{radiation}_{\text{PAR}} \cdot
\big(1-e^{-\,\text{extinction} \cdot \text{leaf area index}}\big)
\cdot \prod_i \text{stress}_i \;-\; \text{respiration}
$$

- **LINTUL** (Light INTerception and UtiLization), Spitters & Schapendonk (1990), potato
  drought. The minimal member of the family: a single radiation use efficiency, and drought
  entering as a multiplier derived from the transpiration ratio.
- **SIMPLE** (Zhao et al. 2019) — deliberately reduced to ~13 parameters; water stress via
  an ARID index.
- **AquaCrop's** canopy-cover formulation is a cousin (it replaces LAI with green canopy
  cover) but its gain term is Backbone C.

Time step: daily. Spatial scale: field / uniform stand. Units: RUE in g DM MJ^-1^ PAR
(typically 2.5–3.5 for C3 grain crops, 3.5–4.5 for C4, on an intercepted-PAR basis).

**Where stress enters**: as a multiplicative dimensionless factor on RUE. This is
computationally trivial but *physiologically wrong in one respect*: RUE is comparatively
conservative under mild drought, because stomatal closure reduces both transpiration and
assimilation roughly in proportion; the first-order drought effect is on **intercepted
radiation** via leaf expansion, not on RUE. Models that put all the drought response on RUE
recover too fast after re-watering, because the leaf area was never lost.

## Backbone B — photosynthesis-driven (light response curve, integrated over canopy)

$$
\text{assimilation}(\text{depth}) = \text{assimilation}_{\max}
\Big(1 - e^{-\,\text{initial slope} \;\cdot\; \text{irradiance}(\text{depth})
\;/\; \text{assimilation}_{\max}}\Big)
$$

integrated over canopy depth and over the day (Gaussian quadrature, typically 3–5 points),
then converted to CH~2~O and reduced by growth and maintenance respiration.

- **SUCROS / WOFOST** — the Wageningen school; de Wit lineage. See van Ittersum et al.
  (2003) for the family overview and de Wit et al. (2019) for 25 years of WOFOST.
- **GECROS** (Yin & van Laar 2005) — same backbone with a much tighter carbon–nitrogen
  coupling and an explicit stomatal/energy-balance layer.
- **Greenhouse models**: reduced **TOMGRO** (Jones et al. 1999) and the
  **Vanthoor** tomato yield model (2011), which is the crop half of a greenhouse
  climate + crop model designed explicitly for control studies.
- **Big-leaf / sun-shade scaling**: de Pury & Farquhar (1997) is the standard route from
  FvCB leaf photosynthesis to a canopy flux without full multi-layer integration.

Time step: daily with sub-daily (hourly or 3-point Gaussian) integration inside.
Spatial scale: canopy layers, one dimension (depth), horizontally uniform.

**Where stress enters**: on the maximum assimilation rate (umol CO~2~ m^-2^ s^-1^) and/or on the stomatal
conductance that sets internal CO~2~. This is more mechanistic and permits a genuine
VPD feedback, at the cost of a stiffer, faster subsystem embedded in a daily model.

## Backbone C — transpiration-driven (normalised water productivity)

$$
\frac{d\,\text{dry mass}}{dt} =
\text{normalised water productivity} \cdot
\frac{\text{transpiration}}{\text{reference evapotranspiration}}
$$

- **AquaCrop** (Steduto et al. 2009 concepts; Raes et al. 2009 algorithms; Hsiao et al.
  2009 maize parameterisation). $\mathrm{WP}^{*}$ is biomass water productivity normalised
  for atmospheric evaporative demand and for CO~2~ concentration, in g DM m^-2^ mm^-1^
  (roughly 15–20 for C3, 30–35 for C4).
- The idea traces to the de Wit (1958) transpiration–production proportionality and to
  Tanner & Sinclair (1983).

**Where stress enters**: nowhere *extra* — it is built in, because transpiration already carries the
water limitation. This is the cleanest formulation for water-limited production and the
reason AquaCrop dominates irrigation-management applications. The price: CO~2~,
temperature and radiation effects have to be smuggled back in through the normalisation, and
the model has almost no purchase on a **salinity** effect that is not mediated by
transpiration (i.e. it handles the osmotic phase well and the ionic-toxicity phase poorly).

## Cross-model contrast worth flagging

The three backbones disagree most sharply about **where a stress factor attaches**:

| | Assimilation rate | Canopy expansion | Senescence | Transpiration |
|---|---|---|---|---|
| LINTUL / SIMPLE | primary | secondary | weak | diagnostic |
| WOFOST / SUCROS | primary (on max assimilation rate) | secondary | via assimilate shortage | computed |
| AquaCrop | via transpiration only | **primary** (Ks~exp~) | **explicit** (Ks~sen~) | primary |
| Vanthoor / TOMGRO | primary + sink limitation | secondary | minor | computed |

AquaCrop's deliberate separation of three stress coefficients — expansion, stomatal,
senescence — each with its own depletion threshold and each with a *different sensitivity
ordering* (expansion most sensitive, senescence least) is, in my view, the single most
useful idea in this literature for anyone building a compact ODE model. It encodes the
time-scale separation directly in the parameter set.

# Water scarcity: from soil potential to stomata

## The macroscopic sink-term tradition (Feddes)

Feddes et al. (1978) reduce root water uptake to a piecewise-linear reduction function
$\text{reduction}(\text{pressure head})$, with four breakpoints: anaerobiosis (too wet), the two
ends of the optimal range, and wilting point. The optimal-range upper breakpoint is made
**demand-dependent**, i.e. it shifts with potential transpiration — an often-overlooked
feedback that makes the "threshold" itself a function of the driver.

- Implemented in **SWAP** (Soil–Water–Atmosphere–Plant; van Dam, Kroes and co-workers) and
  in **HYDRUS-1D** (Simunek et al. 2016).
- Time scale: the sink term is quasi-instantaneous (minutes) but is normally evaluated at
  the soil-solver time step (minutes to hours) inside a daily crop model.
- Spatial scale: one-dimensional soil column, method of lines — i.e. the Richards PDE
  discretised into 10–100 layers, which *is* a stiff ODE system in the frame of this review.

## The hydraulic / signalling tradition

Here stomatal conductance is a state-dependent control rather than a lookup:

- **Tardieu & Davies (1993)**: conductance as a function of xylem ABA concentration *and*
  leaf water potential, with the ABA sensitivity itself modulated by potential. This is the
  cleanest example of a plant-internal $u(t)$ with a perception state (the ABA signal is an
  integrator of root-zone status with a transport lag of hours).
- **Tuzet, Perrier & Leuning (2003)**: a coupled stomatal–photosynthesis–hydraulic model in
  which leaf water potential is solved from a soil-to-leaf resistance network and feeds a
  sigmoidal conductance function. Quasi-steady in water potential (seconds) embedded in a
  daily carbon model.
- **Steppe et al. (2006)**: an explicit ODE for stem water storage and turgor — relevant
  because it makes the capacitance (and hence the diurnal lag between transpiration and sap
  flow, order 1–3 h) a real state rather than an assumption.
- **SurEau / Cochard et al. (2021)**: hydraulic failure and cavitation dynamics; the
  relevant regime for severe scarcity, where the response is no longer reversible.

## The optimal-control tradition

This is the branch that matches the $u(t)$-with-objective framing directly:

- **Cowan & Farquhar (1977)**: stomata maximise carbon gain subject to a water cost,
  $\max \int (\text{assimilation} - \text{water price} \cdot \text{transpiration})\,dt$.
  The Lagrange multiplier "water price" (umol CO~2~ per mol H~2~O) is the marginal
  water-use efficiency and is the natural price of water.
- **Medlyn et al. (2011)**: derives from the same optimisation a unified stomatal model of
  Ball–Berry form, which reconciles the empirical and optimality traditions — a genuinely
  useful cross-check because it shows the empirical slope parameter behaves like the square root of the water price.
- **Sperry et al. (2017)**: carbon-gain maximisation against hydraulic-risk, which
  reproduces the observed behaviour that plants become more conservative as soil dries,
  something a fixed water price cannot do.

**Difference to highlight**: Feddes-type models make water stress a *state-dependent
multiplier*; optimality models make it an *emergent control*. In the frame of this review,
Feddes belongs in $f(\cdot)$ and Cowan–Farquhar belongs in $h_r(\cdot)$ with an explicit
objective. They give similar answers under mild stress and diverge under severe or
intermittent stress, and only the latter can represent a genotype that "chooses" to be
isohydric.

## Time scales (this is the part most compact models get wrong)

| Process | Characteristic time | State it lives in |
|---|---|---|
| Stomatal aperture adjustment | 5–30 min | conductance (often quasi-steady) |
| Leaf/stem capacitance discharge | 1–3 h | tissue water content |
| ABA signal transport root to leaf | 2–12 h | xylem ABA concentration |
| Osmotic adjustment (solute accumulation) | 1–7 d | tissue osmolality |
| Leaf expansion suppression | hours to days | leaf area index |
| Root:shoot allocation shift | 5–20 d | partitioning fractions |
| Accelerated senescence / leaf shedding | 1–3 weeks | leaf area index, leaf sodium |
| Root-zone salt accumulation | 2 weeks – 1 season | root-zone salt |
| Sodicity / structural degradation | 1–10 seasons | exchangeable sodium percentage |
| Soil pH change (mineral soil) | 1–10 years | base saturation, alkalinity |
| Substrate pH change (rockwool, coir) | hours – days | solution alkalinity |

A model that puts salinity stress on the same time scale as stomatal closure will
systematically mispredict recovery after a leaching irrigation.

# Salinity: osmotic phase, ionic phase, and how they are combined

## The two-phase conceptual model

Munns (1993) and Munns & Tester (2008) are the reference framing: a **fast osmotic phase**
(within minutes to days of salt reaching the root, indistinguishable from drought in its
effect on expansion and conductance) followed by a **slow ionic phase** (weeks; Na^+^ and
Cl^-^ accumulate in old leaves beyond the compartmentation capacity, causing premature
senescence and therefore loss of assimilating area).

Implication for ODE structure, and this is important: **the two phases need different
states**. The osmotic phase is a function of a driver-like quantity (root-zone EC now); the
ionic phase requires an *integrator* state — leaf tissue sodium, whose derivative is
transpiration times solution concentration times a bypass fraction. Almost no
production-scale crop model carries this second state; AquaCrop, WOFOST and DSSAT do not.
This is the clearest gap in the field-scale literature.

## Static response functions (still the workhorse)

- **Maas & Hoffman (1977)**: relative yield as threshold-and-slope in EC of the saturated
  paste extract, $\text{relative yield} = 1 - \text{slope} \cdot
  (\mathrm{EC}_e - \mathrm{EC}_{\text{threshold}})$. Still the
  basis of FAO-29 tables. Piecewise-linear, hence non-differentiable — replace with a
  smoothed version before putting it in a fixed-step integrator.
- **van Genuchten & Hoffman (1984)** / van Genuchten (1987): the S-shaped alternative,
  $\text{relative yield} = 1/\big(1 + (\mathrm{EC}/\mathrm{EC}_{50})^{\text{shape}}\big)$,
  with the shape exponent typically 3. Smooth,
  one fewer parameter, and better behaved near the threshold.
- Both are **steady-state, whole-season** relationships. Using them as an instantaneous
  reduction factor inside a dynamic model is a category error that the literature commits
  routinely; Letey & Feng (2007) is the explicit critique (dynamic versus steady-state
  salinity–yield modelling, and why the steady-state leaching requirement overestimates the
  water needed).

## Combining matric and osmotic stress: three incompatible formulations

This is the point where models genuinely disagree, and it deserves the sharpest statement
in this review.

1. **Multiplicative**: $\text{reduction}_{\text{total}} =
   \text{reduction}(\text{matric}) \cdot \text{reduction}(\text{osmotic})$.
   Used in HYDRUS and SWAP by default. Assumes independence.
2. **Additive in potential**: $\text{reduction}_{\text{total}} =
   \text{reduction}(\text{matric} + \text{osmotic})$
   (van Genuchten 1987). Assumes the plant responds to total water potential and cannot tell
   the two apart.
3. **Non-equivalent / weighted**: matric and osmotic potentials are *not* interchangeable,
   because (a) osmotic potential is spatially uniform in the root zone while matric potential
   develops steep gradients at the root surface, and (b) salinity itself reduces root
   hydraulic conductance. Shani & Dudley (2001) and Shani et al. (2007) build an analytical
   model integrating yield, water, **soil type** and salinity on this basis; Homaee, Feddes
   & Dirksen (2002) test the alternatives and find the multiplicative form biased.
   Skaggs et al. (2006) is the systematic comparison of macroscopic approaches.

**My reading of the disagreement**: the multiplicative form is conservative (predicts more
stress) and is defensible when matric stress is mild and salinity dominates — the typical
drip-irrigated or soilless case. The additive form is defensible in dry-farmed saline soils
where both potentials are large. The Shani/Ben-Gal line is the only one that gets the
*soil-type interaction* right, because it recognises that in a coarse sand the matric
component collapses quickly between irrigations while in a clay it does not, so the same
seasonal EC produces different dynamics. If your question is about salinity build-up under
irrigation management, use formulation 3 or at minimum make the choice a switch and report
sensitivity to it — which is what the code backbone below does.

## The salt balance and the build-up mechanism

The build-up itself is a two-line ODE and is not controversial:

$$
\frac{d(\text{root-zone salt})}{dt} =
  \text{conc}_{\text{irrigation}} \cdot \text{irrigation}
+ \text{conc}_{\text{rain}} \cdot \text{rainfall}
- \text{conc}_{\text{soil}} \cdot \text{drainage}
- \text{uptake fraction} \cdot \text{conc}_{\text{soil}} \cdot \text{transpiration}
$$

with the salt uptake fraction (one minus the exclusion fraction) typically 0.02–0.10 for
glycophyte crops.
**The mechanism is simply that transpiration removes water and leaves salt behind**: a crop
transpiring 500 mm of water at 1.0 dS m^-1^ imports roughly 320 g m^-2^ of salt in a season
and exports only a few percent of it in the harvest.

At steady state with a leaching fraction defined as drainage divided by applied water, the
classical result is $\text{conc}_{\text{drain}} = \text{conc}_{\text{irrigation}} /
\text{leaching fraction}$, i.e. drain EC is inversely proportional to
leaching fraction (Rhoades; the FAO-29 leaching-requirement framework). The transient to
that steady state takes weeks to months in soil and **days** in a low-volume substrate,
which is exactly why the greenhouse literature had to develop its own dynamic version.

## Sodicity and the soil-type coupling (EC is not only an osmotic driver)

Low EC combined with a high sodium adsorption ratio (SAR) causes clay swelling and
dispersion, collapsing saturated hydraulic conductivity — so *reducing* EC by leaching with
low-salinity water can make infiltration worse. The threshold electrolyte concentration
(TEC) concept is from Quirk & Schofield (1955); the quantitative model is McNeal (1968),
modernised by Ezlit et al. (2013). This coupling exists only in 2:1 clay soils
(smectite/vermiculite dominated), is weak in kaolinitic and sandy soils, and is absent in
inert substrates (rockwool, perlite) — which is the main reason soilless and field salinity
models are not interchangeable.

# EC and salinity build-up in soilless / greenhouse systems

This is where the dynamics are fastest and the control is tightest, and where the ODE
formulations are genuinely at their best.

- **Sonneveld & Voogt (2009)**, *Plant Nutrition of Greenhouse Crops* — the conceptual
  anchor: the **uptake concentration** $C_{\text{uptake}}$ (mmol per litre of water taken
  up) for each ion. If supply concentration exceeds uptake concentration, that ion
  accumulates; Na^+^ and Cl^-^ have near-zero uptake concentration in most crops, so they
  accumulate fastest. This single concept explains virtually all observed EC drift.
- **Carmassi et al. (2005, 2007)**: an explicit dynamic model of salinity build-up in
  recirculating (closed) nutrient solution culture for greenhouse tomato, including the
  discharge/refresh decision. Time scale: days; state: solution volume and per-ion mass;
  control: discharge fraction and refresh EC.
- **Massa, Incrocci, Pardossi and co-workers (2011)**: strategies to reduce drainage volume
  and nitrate emission from soilless tomato — the same model used as a control-design tool,
  with an explicit multi-objective trade-off (yield versus emission).
- **Vanthoor et al. (2011)** plus the companion greenhouse climate model: the standard
  crop-side backbone for control studies; the crop model is Backbone B with sink limitation.
- **van Straten, van Willigenburg, van Henten & van Ooteghem (2010)**, *Optimal Control of
  Greenhouse Cultivation* (CRC Press): the reference text for exactly the $u(t)$-with-
  objective-function framing. Two-time-scale decomposition (fast climate, slow crop) with a
  receding-horizon outer loop is the method of choice here.

**Key structural difference versus field models**: in a substrate the root-zone water volume
is small (rockwool slab: 6–10 L m^-2^; typical field root zone: 100–200 mm of available
water = 100–200 L m^-2^), so the salt-balance ODE has a time constant of days rather than
months, and the drain EC is measurable continuously. That makes closed-loop EC control
feasible, and makes the drain EC an excellent *observable* $Y(t)$ for state estimation of
the root-zone salt state — which the field case lacks entirely.

# pH and alkalinity

pH is the weakest-developed channel in the crop-model literature, and honesty requires
saying so: in WOFOST, AquaCrop, CERES/DSSAT and STICS, pH does **not** affect assimilation
at all. Where it does appear:

- **EPIC / APEX** (Williams and co-workers): carries an **aluminium saturation stress
  factor** that reduces crop growth in acid soils, and a pH-dependent phosphorus sorption
  model. This is the main field-scale precedent for a genuine pH-to-growth pathway.
- **UNSATCHEM** (Simunek & Suarez 1994; Suarez & Simunek 1997), available inside the HYDRUS
  family: solves major-ion chemistry, CO~2~ production and transport, calcite
  precipitation/dissolution, cation exchange and gypsum dynamics, and therefore computes pH,
  EC and SAR as *states* rather than drivers. This is the most complete treatment I am aware
  of at the profile scale.
- **Greenhouse / soilless**: pH is a fast state governed by an alkalinity balance —
  bicarbonate input from irrigation water, acid dosing (a genuine control), and
  plant-driven proton excretion set by the cation:anion uptake ratio, which is dominated by
  the ammonium:nitrate ratio of the supply. Nitrate-fed crops raise solution pH; ammonium-fed
  crops lower it. Buffer capacity of the root environment differs by three orders of
  magnitude between rockwool (essentially none) and a calcareous clay loam.
- **Rhizosphere pH gradients**: Nye (1981) and later reactive-transport treatments; relevant
  at the millimetre scale, and mostly out of scope here except as the mechanism behind the
  bulk effect.

**Mechanistic pathway to assimilation**, when it is represented at all, is indirect and
slow: pH controls micronutrient availability (Fe, Mn, Zn); Fe deficiency reduces chlorophyll
and therefore the maximum assimilation rate on a time scale of 1–3 weeks; at low pH, Al^3+^ toxicity restricts
root elongation and therefore the explored soil volume, which feeds back into water and
nutrient capture. Both are best modelled as slow multipliers on the maximum assimilation rate (or RUE) and on
effective rooting depth, not as instantaneous stress factors.

# Soil and substrate type: where it actually enters

Four distinct entry points, often conflated:

1. **Water retention and conductivity** — van Genuchten (1980) parameters; sets how quickly
   matric potential falls between irrigations. Sand: steep, fast; clay: shallow, slow. In a
   lumped ODE this shows up as the exponent of the retention curve and as the available
   water capacity (mm per m of profile: coarse sand 60–100, loam 150–200, clay loam 160–220,
   rockwool effectively a bucket with near-zero tension range).
2. **Salt storage and buffering** — the same water-holding capacity sets the salt-balance
   time constant. A soil with more water holds more salt at the same EC and responds more
   slowly to a change in irrigation EC.
3. **Chemical buffering** — cation exchange capacity and carbonate content set the pH
   buffer capacity and the SAR–ESP relationship; also gypsum dissolution as a natural EC
   floor in gypsiferous soils.
4. **Structural response to EC/SAR** — the TEC/dispersion coupling above, present only in
   2:1 clays.

Only the Shani/Ben-Gal analytical line and the HYDRUS/UNSATCHEM family represent all four.
Most crop models represent only (1).

# Steering: controls, policies and objectives

## Controls that appear in this literature

| Control | Units | Time scale of decision | Where it acts |
|---|---|---|---|
| Irrigation amount | mm per event | hours – days | water balance |
| Irrigation frequency / "meal pattern" | events d^-1^ | days | matric potential amplitude |
| Target leaching fraction | – | season, adjusted weekly | salt balance |
| Supply EC set point | dS m^-1^ | days | osmotic potential, fruit quality |
| Drain/discharge fraction (soilless) | – | days | salt build-up |
| Acid or base dosing | mmol~c~ m^-2^ d^-1^ | hours | alkalinity, pH |
| Nitrate:ammonium ratio | – | days | proton balance, pH drift |
| Deficit strategies: RDI, PRD | – | phenological stage | stomatal and expansion stress |
| CO~2~ enrichment, screens, ventilation | umol mol^-1^, – | minutes | assimilation, VPD |

## Rule-based policies with perception states

The policies actually used are rule-based, not optimal, and they carry their own dynamics:

- **Soil-moisture-triggered irrigation**: a tensiometer or capacitance sensor with a
  first-order lag and a dead band. The lag constant (0.2–2 d, depending on sensor placement
  and soil contact) is a genuine state, and it matters: a slow sensor in a coarse substrate
  produces oscillatory over-irrigation.
- **Drain-EC feedback**: increase leaching fraction when measured drain EC exceeds a set
  point. This is the standard salinity policy in greenhouse horticulture and is a
  proportional controller on a slow, well-observed state.
- **Radiation-sum triggered irrigation**: the dominant rule in Dutch greenhouse practice —
  irrigate after each accumulated radiation increment (e.g. every 100–150 J cm^-2^). This
  is a *feedforward* controller on $w(t)$ with an integrator state, and it is remarkably
  robust precisely because it anticipates transpiration rather than reacting to depletion.

In the frame of this review all three are $h_r(X, p_u, w)$ with their own states appended to
$X$, and $p_u$ (trigger thresholds, gains, lag constants, set points) is what gets optimised.

## Objective functions used

- Maximum yield subject to a water quota (the classical deficit-irrigation problem).
- Maximum gross margin: yield value minus water, energy, fertiliser and disposal cost.
- Multi-objective with an emission constraint (nitrate leaching), as in the Massa/Incrocci
  work.
- Cowan–Farquhar marginal water-use efficiency at the plant level, which is the same
  problem one level down, and a useful consistency check: an irrigation policy that implies
  a water price far from the plant's own is likely mis-specified.
- Long-horizon sustainability: minimise root-zone salt at end of season subject to a yield
  floor. This is the objective that makes the salt state worth carrying at all, since within
  a single season leaching is almost always a net cost.

# Model comparison table

| Model | Backbone | Time step | Spatial scale | Water stress route | Salinity | pH | Soil/substrate type | Main control |
|---|---|---|---|---|---|---|---|---|
| LINTUL | A | day | field | RUE multiplier | none | none | bucket | irrigation |
| WOFOST / SUCROS | B | day (Gaussian intra-day) | field, canopy layers | max assimilation rate + expansion | none native (via SWAP coupling) | none | retention curve | irrigation, sowing date |
| AquaCrop | C | day | field | Ks~exp~, Ks~sto~, Ks~sen~ | Maas–Hoffman-type Ks~salt~ | none | AWC + drainage | irrigation, LF |
| DSSAT / CERES | A/B hybrid | day | field | supply:demand ratio | limited | none | layered bucket | irrigation, N |
| APSIM | A/B hybrid | day | field, layered | supply:demand + root `kl` | via root exploration factor | indirect | layered, well developed | irrigation, N, rotation |
| STICS | A | day | field | turgor + stomatal | limited | none | layered | irrigation, N |
| CropSyst | A | day | field | transpiration ratio | salinity module | none | layered | irrigation |
| SWAP (+WOFOST) | B | day / soil sub-step | 1-D column | Feddes sink term | osmotic head, multiplicative | none | full van Genuchten | irrigation, drainage |
| HYDRUS-1D | – (transport) | adaptive, minutes | 1-D/2-D | Feddes + osmotic | full transport | via UNSATCHEM | full | boundary flux |
| UNSATCHEM | – | adaptive | 1-D | Feddes | full major-ion chemistry | **explicit state** | full + exchange | water quality, amendments |
| SALTMED | A/C | day | field, 2-D-ish | reduction function | explicit, integrated | limited | layered | irrigation, LF, fertigation |
| Shani et al. (2007) | analytical | season | root zone | combined potential | non-equivalent combination | none | **explicit soil type** | irrigation, EC |
| TOMGRO / Vanthoor | B | hour–day | greenhouse compartment | via conductance/VPD | via EC set point | via solution | inert substrate | climate + fertigation |
| Carmassi et al. | mass balance | day | slab / loop | – | **build-up as state** | solution | inert substrate | discharge fraction |
| EPIC / APEX | A | day | field | water stress factor | salinity factor | **Al saturation** | full | management |
| SahysMod | balance | season | polygon / region | – | regional salt + water table | – | aggregated | drainage, LF |

# Model backbone as R code

Deliberately compact: one lumped root zone, one crop, all four channels (water, EC, salt
build-up, pH) present, and the irrigation/EC/pH policy inside the same state vector.
Parameter *values* are placeholders, not calibrations — the point is the structure, the
units and the switches that let the formulation disagreements above be tested.

Every threshold is smoothed, because the integration is fixed-step Runge–Kutta 4.

## Helper functions

```{r}
#| label: helpers

library(deSolve)
library(data.table)
library(ggplot2)

# --------------------------------------------------------------------------
# Smooth replacements for min / max / step.
# Fixed-step rk4 needs a continuously differentiable right-hand side; a hard
# min() or an if() on a state creates a discontinuity that rk4 will silently
# integrate across with an O(1) error.
# --------------------------------------------------------------------------

soft_min <- function(value_a, value_b, smoothing = 1e-3) {
  # value_a, value_b [any, same units]; smoothing [same units] = corner radius
  0.5 * (value_a + value_b - sqrt((value_a - value_b)^2 + smoothing^2))
}

soft_max <- function(value_a, value_b, smoothing = 1e-3) {
  # value_a, value_b [any, same units]; smoothing [same units]
  0.5 * (value_a + value_b + sqrt((value_a - value_b)^2 + smoothing^2))
}

soft_step <- function(value, threshold, steepness) {
  # value, threshold [any, same units]; steepness [1 / units of value]
  # returns [-] in (0, 1)
  1 / (1 + exp(-steepness * (value - threshold)))
}

bell_factor <- function(value, optimum, width) {
  # symmetric response, e.g. pH optimum; returns [-] in (0, 1]
  exp(-((value - optimum) / width)^2)
}
```

## The ODE system

```{r}
#| label: model-backbone

crop_water_salinity_model <- function(time, state, parms) {

  # time  [d] since emergence
  # state named numeric vector, order must match the derivative vector returned
  # parms list holding parameters, initial states and driver functions

  list2env(c(parms, as.list(state)), envir = environment())

  # ------------------------------------------------------------------------
  # 1. Uncontrolled drivers w(time)
  # ------------------------------------------------------------------------
  global_radiation      <- driver_global_radiation(time)       # [MJ m-2 d-1] total shortwave
  reference_et          <- driver_reference_et(time)           # [mm d-1] FAO-56 ET0
  air_temperature       <- driver_air_temperature(time)        # [degC] daily mean
  rainfall              <- driver_rainfall(time)               # [mm d-1]
  irrigation_water_ec   <- driver_irrigation_water_ec(time)    # [dS m-1]
  irrigation_alkalinity <- driver_irrigation_alkalinity(time)  # [mmol_c L-1] as HCO3-

  # ------------------------------------------------------------------------
  # 2. Root-zone status (algebraic, from states)
  # ------------------------------------------------------------------------
  water_capacity <- root_zone_depth * available_water_capacity
  # root_zone_depth [m]; available_water_capacity [mm m-1] -> water_capacity [mm]

  relative_water <- root_zone_water / water_capacity                    # [-] 0..1
  water_deficit  <- 1 - relative_water                                  # [-]

  solution_volume <- root_zone_water + water_volume_floor               # [mm] = [L m-2]
  salt_concentration <- root_zone_salt / solution_volume                # [g L-1]
  soil_ec <- salt_concentration / salt_per_ec                           # [dS m-1]
  # salt_per_ec [g L-1 per dS m-1], about 0.64 for a mixed-salt solution

  sodium_concentration <- soil_ec * charge_per_ec * sodium_fraction     # [mmol_c L-1]
  # charge_per_ec [mmol_c L-1 per dS m-1], about 10

  osmotic_potential <- -osmotic_per_ec * soil_ec                        # [MPa], negative
  # osmotic_per_ec [MPa per dS m-1], about 0.036 (van 't Hoff, Richards 1954)

  matric_potential <- matric_potential_at_capacity *
    (relative_water + relative_water_floor)^(-retention_exponent)       # [MPa], negative
  # matric_potential_at_capacity [MPa] about -0.01
  # retention_exponent [-] about 4 for sand, 6 loam, 9 clay  <- SOIL TYPE ENTERS HERE

  total_potential <- matric_potential + osmotic_potential               # [MPa]

  # ------------------------------------------------------------------------
  # 3. Stress factors  -- the three literature formulations as a switch
  # ------------------------------------------------------------------------
  # Individual factors, both S-shaped (van Genuchten 1987 form, smoothed
  # Maas-Hoffman), all dimensionless in (0, 1]
  stress_water_stomatal <- 1 / (1 + (water_deficit / deficit_50_stomatal)^stress_shape)
  stress_water_expansion <- 1 / (1 + (water_deficit / deficit_50_expansion)^stress_shape)
  stress_salt <- 1 / (1 + (soil_ec / ec_50)^salinity_shape)
  stress_potential <- 1 / (1 + (abs(total_potential) / potential_50)^stress_shape)

  if (stress_combination == "multiplicative") {
    # HYDRUS / SWAP default: independence of matric and osmotic stress
    stress_stomatal  <- stress_water_stomatal * stress_salt
    stress_expansion <- stress_water_expansion * stress_salt
  } else if (stress_combination == "additive_potential") {
    # van Genuchten (1987): plant responds to total water potential only
    stress_stomatal  <- stress_potential
    stress_expansion <- stress_potential^expansion_sensitivity_ratio
  } else {
    # Non-equivalent / limiting-factor (Shani, Homaee): smoothed minimum,
    # with osmotic stress additionally reducing root hydraulic conductance
    root_conductance_factor <- 1 / (1 + (soil_ec / ec_50_conductance)^salinity_shape)
    stress_stomatal  <- soft_min(value_a = stress_water_stomatal,
                                 value_b = stress_salt, smoothing = 0.02) *
      root_conductance_factor
    stress_expansion <- soft_min(value_a = stress_water_expansion,
                                 value_b = stress_salt, smoothing = 0.02)
  }

  # Ionic (Munns phase 2): slow, integrator-driven, acts on senescence not on
  # instantaneous assimilation
  leaf_water_volume <- leaf_area_index * leaf_water_per_area + water_volume_floor  # [L m-2]
  leaf_sodium_concentration <- leaf_sodium / leaf_water_volume          # [mmol_c L-1]

  # ------------------------------------------------------------------------
  # 4. Canopy, water fluxes
  # ------------------------------------------------------------------------
  cover_fraction <- 1 - exp(-extinction_coefficient * leaf_area_index)  # [-]
  # extinction_coefficient [m2 ground m-2 leaf], about 0.6-0.8

  potential_transpiration <- reference_et * crop_coefficient_basal * cover_fraction  # [mm d-1]
  actual_transpiration    <- potential_transpiration * stress_stomatal              # [mm d-1]
  soil_evaporation <- reference_et * (1 - cover_fraction) * evaporation_factor      # [mm d-1]

  # ------------------------------------------------------------------------
  # 5. Control u(time) = h_r(X, p_u, w) -- rule based, with perception states
  # ------------------------------------------------------------------------
  # Leaching fraction: base value, raised proportionally when the PERCEIVED
  # drain EC exceeds its set point (standard greenhouse salinity policy)
  leaching_fraction <- soft_min(
    value_a = base_leaching_fraction +
      leaching_gain * soft_step(value = perceived_ec, threshold = ec_set_point,
                                steepness = ec_steepness),
    value_b = maximum_leaching_fraction, smoothing = 0.01)              # [-]

  # Irrigation: triggered on perceived depletion, sized to replace transpiration
  # and deliver the target leaching fraction, capped by system capacity
  irrigation_trigger <- soft_step(value = perceived_deficit, threshold = deficit_trigger,
                                  steepness = trigger_steepness)        # [-]
  irrigation_rate <- soft_min(
    value_a = irrigation_trigger * potential_transpiration / (1 - leaching_fraction),
    value_b = maximum_irrigation_rate, smoothing = 0.05)                # [mm d-1]

  # Acid dosing: proportional control on substrate pH above its set point
  acid_dose <- acid_gain * soft_max(value_a = substrate_ph - ph_set_point,
                                    value_b = 0, smoothing = 0.02)      # [mmol_c m-2 d-1]

  # ------------------------------------------------------------------------
  # 6. Assimilation -- von Bertalanffy structure:
  #    surface-like gain (light capture saturates with leaf area)
  #    minus mass-like loss (maintenance proportional to standing biomass)
  # ------------------------------------------------------------------------
  temperature_factor <- bell_factor(value = air_temperature, optimum = temperature_optimum,
                                    width = temperature_width)          # [-]
  micronutrient_factor <- bell_factor(value = substrate_ph, optimum = ph_optimum,
                                      width = ph_width)                 # [-]
  # pH acts slowly and indirectly (Fe/Mn availability -> chlorophyll -> Amax);
  # applied here as a multiplier on the capacity term, not on conductance.

  if (assimilation_mode == "radiation") {
    # Backbone A: LINTUL / SIMPLE
    intercepted_radiation <- global_radiation * par_fraction * cover_fraction  # [MJ PAR m-2 d-1]
    supply_rate <- radiation_use_efficiency * intercepted_radiation *
      temperature_factor * micronutrient_factor * stress_stomatal       # [g DM m-2 d-1]
  } else {
    # Backbone C: AquaCrop, transpiration driven
    supply_rate <- normalised_water_productivity *
      (actual_transpiration / (reference_et + reference_et_floor)) *
      temperature_factor * micronutrient_factor                         # [g DM m-2 d-1]
  }

  # Sink capacity: temperature driven demand, saturating with existing mass
  demand_rate <- maximum_sink_rate * temperature_factor *
    crop_dry_mass / (crop_dry_mass + sink_half_saturation)              # [g DM m-2 d-1]

  gross_assimilation <- soft_min(value_a = supply_rate, value_b = demand_rate,
                                 smoothing = 0.2)  # the a() gate [g DM m-2 d-1]

  maintenance_respiration <- maintenance_coefficient * crop_dry_mass *
    q_ten^((air_temperature - reference_temperature) / 10)              # [g DM m-2 d-1]

  net_growth <- soft_max(value_a = gross_assimilation - maintenance_respiration,
                            value_b = 0, smoothing = 0.05)

  # ------------------------------------------------------------------------
  # 7. Derivatives
  # ------------------------------------------------------------------------
  senescence_rate <- (base_senescence_rate +
    sodium_senescence_coefficient *
      soft_max(value_a = leaf_sodium_concentration - sodium_tolerance,
               value_b = 0, smoothing = 1)) *
    leaf_area_index                                                     # [m2 m-2 d-1]

  d_crop_dry_mass <- net_growth                                         # [g DM m-2 d-1]

  d_leaf_area_index <- specific_leaf_area * leaf_fraction * net_growth *
    stress_expansion - senescence_rate                                  # [m2 m-2 d-1]

  drainage <- drainage_rate_constant *
    soft_max(value_a = root_zone_water - water_capacity, value_b = 0,
             smoothing = 0.5)                                           # [mm d-1]

  d_root_zone_water <- irrigation_rate + rainfall - actual_transpiration -
    soil_evaporation - drainage                                         # [mm d-1]

  salt_input  <- irrigation_rate * irrigation_water_ec * salt_per_ec    # [g m-2 d-1]
  salt_drained <- drainage * salt_concentration                         # [g m-2 d-1]
  salt_taken_up <- actual_transpiration * salt_concentration * salt_uptake_fraction
  d_root_zone_salt <- salt_input - salt_drained - salt_taken_up         # [g m-2 d-1]

  d_leaf_sodium <- actual_transpiration * sodium_concentration * sodium_bypass_fraction -
    leaf_sodium * senescence_rate / (leaf_area_index + leaf_area_floor) # [mmol_c m-2 d-1]

  # Alkalinity balance expressed directly as a pH rate through a buffer capacity
  proton_excretion <- proton_per_growth * net_growth *
    (2 * ammonium_fraction - 1)                                         # [mmol_c m-2 d-1]
  alkalinity_input   <- irrigation_rate * irrigation_alkalinity         # [mmol_c m-2 d-1]
  alkalinity_drained <- drainage * irrigation_alkalinity * alkalinity_retention
  d_substrate_ph <- (alkalinity_input - alkalinity_drained - acid_dose -
    proton_excretion) / buffer_capacity                                 # [pH unit d-1]
  # buffer_capacity [mmol_c m-2 per pH unit]: rockwool ~5, sandy soil ~200,
  # calcareous clay loam ~5000   <- SOIL / SUBSTRATE TYPE ENTERS HERE AS WELL

  # Perception states of the controller (first order sensor lag)
  d_perceived_deficit <- (water_deficit - perceived_deficit) / sensor_time_constant_water
  d_perceived_ec      <- (soil_ec - perceived_ec) / sensor_time_constant_ec

  d_cumulative_transpiration <- actual_transpiration                    # [mm d-1]
  d_cumulative_irrigation    <- irrigation_rate                         # [mm d-1]
  d_cumulative_drainage      <- drainage                                # [mm d-1]

  # ------------------------------------------------------------------------
  # 8. Outputs Y(time) = g(X, p, u, w)
  # ------------------------------------------------------------------------
  list(
    c(d_crop_dry_mass,
      d_leaf_area_index,
      d_root_zone_water,
      d_root_zone_salt,
      d_leaf_sodium,
      d_substrate_ph,
      d_perceived_deficit,
      d_perceived_ec,
      d_cumulative_transpiration,
      d_cumulative_irrigation,
      d_cumulative_drainage),
    soil_ec                   = soil_ec,                    # [dS m-1]
    drain_ec                  = salt_concentration / salt_per_ec,  # [dS m-1]
    osmotic_potential         = osmotic_potential,          # [MPa]
    matric_potential          = matric_potential,           # [MPa]
    stress_stomatal           = stress_stomatal,            # [-]
    stress_expansion          = stress_expansion,           # [-]
    stress_salt               = stress_salt,                # [-]
    leaf_sodium_concentration = leaf_sodium_concentration,  # [mmol_c L-1]
    cover_fraction            = cover_fraction,             # [-]
    actual_transpiration      = actual_transpiration,       # [mm d-1]
    potential_transpiration   = potential_transpiration,    # [mm d-1]
    irrigation_rate           = irrigation_rate,            # [mm d-1]
    leaching_fraction         = leaching_fraction,          # [-]
    drainage                  = drainage,                   # [mm d-1]
    acid_dose                 = acid_dose,                  # [mmol_c m-2 d-1]
    gross_assimilation        = gross_assimilation,         # [g DM m-2 d-1]
    maintenance_respiration   = maintenance_respiration,    # [g DM m-2 d-1]
    harvestable_yield         = harvest_index * crop_dry_mass  # [g DM m-2]
  )
}
```

## Parameters, drivers and initial states

```{r}
#| label: parameters

make_drivers <- function(season_length = 150, random_seed = 1) {
  # season_length [d]; returns a list of approxfun objects, one per driver
  set.seed(random_seed)
  day_index <- 0:season_length                                          # [d]

  number_of_days <- length(day_index)                                   # [-]
  radiation <- 12 + 8 * sin(2 * pi * (day_index - 40) / 365) +
    rnorm(n = number_of_days, mean = 0, sd = 2)                         # [MJ m-2 d-1]
  temperature <- 16 + 8 * sin(2 * pi * (day_index - 50) / 365) +
    rnorm(n = number_of_days, mean = 0, sd = 1.5)                       # [degC]
  reference_et <- pmax(0.3, 0.35 * radiation * (1 + 0.02 * (temperature - 16)))  # [mm d-1]
  rainfall <- rbinom(n = number_of_days, size = 1, prob = 0.12) *
    rexp(n = number_of_days, rate = 1 / 8)                              # [mm d-1]

  list(
    driver_global_radiation = approxfun(x = day_index, y = pmax(radiation, 0.5), rule = 2),
    driver_reference_et     = approxfun(x = day_index, y = reference_et, rule = 2),
    driver_air_temperature  = approxfun(x = day_index, y = temperature, rule = 2),
    driver_rainfall         = approxfun(x = day_index, y = rainfall, rule = 2),
    driver_irrigation_water_ec = approxfun(
      x = day_index, y = rep(1.2, number_of_days), rule = 2),           # [dS m-1]
    driver_irrigation_alkalinity = approxfun(
      x = day_index, y = rep(4.0, number_of_days), rule = 2)            # [mmol_c L-1]
  )
}

make_parms <- function(drivers,
                       stress_combination = "non_equivalent",
                       assimilation_mode  = "radiation",
                       base_leaching_fraction = 0.20,
                       ec_set_point = 3.0) {

  # stress_combination: "multiplicative" | "additive_potential" | "non_equivalent"
  # assimilation_mode : "radiation" (Backbone A) | "transpiration" (Backbone C)

  parms <- list(

    # ---- switches -------------------------------------------------------
    stress_combination = stress_combination,
    assimilation_mode  = assimilation_mode,

    # ---- canopy and carbon ---------------------------------------------
    extinction_coefficient   = 0.70,   # [m2 ground m-2 leaf]
    par_fraction             = 0.50,   # [MJ PAR per MJ global]
    radiation_use_efficiency = 3.0,    # [g DM MJ-1 intercepted PAR]
    normalised_water_productivity = 18,# [g DM m-2 mm-1] AquaCrop WP*
    specific_leaf_area       = 0.022,  # [m2 leaf g-1 leaf DM]
    leaf_fraction            = 0.45,   # [g leaf DM per g DM]
    harvest_index            = 0.45,   # [-]
    maximum_sink_rate        = 32,     # [g DM m-2 d-1]
    sink_half_saturation     = 40,     # [g DM m-2]
    maintenance_coefficient  = 0.012,  # [g DM g-1 DM d-1] at reference temperature
    q_ten                    = 2.0,    # [-]
    reference_temperature    = 20,     # [degC]
    temperature_optimum      = 22,     # [degC]
    temperature_width        = 12,     # [degC]
    base_senescence_rate     = 0.010,  # [d-1]

    # ---- water and soil / substrate type -------------------------------
    root_zone_depth            = 0.50, # [m]
    available_water_capacity   = 180,  # [mm m-1]  loam
    retention_exponent         = 6.0,  # [-] 4 sand, 6 loam, 9 clay
    matric_potential_at_capacity = -0.010, # [MPa]
    crop_coefficient_basal     = 1.05, # [-]
    evaporation_factor         = 0.35, # [-]
    drainage_rate_constant     = 1.5,  # [d-1]

    # ---- salinity -------------------------------------------------------
    salt_per_ec         = 0.64,  # [g L-1 per dS m-1]
    charge_per_ec       = 10.0,  # [mmol_c L-1 per dS m-1]
    osmotic_per_ec      = 0.036, # [MPa per dS m-1]
    sodium_fraction     = 0.40,  # [mol_c Na per mol_c total]
    salt_uptake_fraction = 0.05, # [-] exclusion: 95 % of salt stays behind
    sodium_bypass_fraction = 0.02, # [-] apoplastic bypass to the shoot
    ec_50               = 6.0,   # [dS m-1] 50 % reduction point
    ec_50_conductance   = 9.0,   # [dS m-1] root hydraulic conductance
    salinity_shape      = 3.0,   # [-] van Genuchten exponent
    sodium_tolerance    = 60,    # [mmol_c L-1] leaf sap
    sodium_senescence_coefficient = 2.0e-4, # [d-1 per mmol_c L-1]
    leaf_water_per_area = 0.20,  # [L m-2 leaf]

    # ---- water stress shape --------------------------------------------
    deficit_50_stomatal  = 0.55, # [-] fraction of available water depleted
    deficit_50_expansion = 0.35, # [-] expansion is more sensitive (AquaCrop)
    stress_shape         = 4.0,  # [-]
    potential_50         = 0.60, # [MPa]
    expansion_sensitivity_ratio = 1.6, # [-]

    # ---- pH and alkalinity ---------------------------------------------
    buffer_capacity      = 200,  # [mmol_c m-2 per pH unit]
    ph_optimum           = 6.0,  # [-]
    ph_width             = 1.2,  # [pH unit]
    proton_per_growth    = 0.9,  # [mmol_c per g DM]
    ammonium_fraction    = 0.10, # [mol N-NH4 per mol N total]
    alkalinity_retention = 0.60, # [-]

    # ---- controller p_u (this is what gets optimised) -------------------
    deficit_trigger          = 0.35, # [-]
    trigger_steepness        = 25,   # [-]
    base_leaching_fraction   = base_leaching_fraction, # [-]
    leaching_gain            = 0.25, # [-]
    maximum_leaching_fraction = 0.55,# [-]
    ec_set_point             = ec_set_point, # [dS m-1]
    ec_steepness             = 2.5,  # [per dS m-1]
    maximum_irrigation_rate  = 12,   # [mm d-1]
    ph_set_point             = 6.2,  # [-]
    acid_gain                = 60,   # [mmol_c m-2 d-1 per pH unit]
    sensor_time_constant_water = 0.5,# [d]
    sensor_time_constant_ec    = 2.0,# [d]

    # ---- numerical floors ----------------------------------------------
    water_volume_floor   = 1.0,   # [mm]
    relative_water_floor = 0.02,  # [-]
    reference_et_floor   = 0.05,  # [mm d-1]
    leaf_area_floor      = 0.01,  # [m2 m-2]

    # ---- initial states (kept in parms, per house style) ----------------
    initial_crop_dry_mass    = 8,    # [g DM m-2]
    initial_leaf_area_index  = 0.15, # [m2 m-2]
    initial_root_zone_water  = 76.5, # [mm] = 0.85 * capacity
    initial_root_zone_salt   = 61.4, # [mm * dS m-1 * g L-1 dS-1 m] -> [g m-2], EC ~ 1.25
    initial_leaf_sodium      = 0,    # [mmol_c m-2]
    initial_substrate_ph     = 6.0,  # [-]
    initial_perceived_deficit = 0.15,# [-]
    initial_perceived_ec     = 1.25, # [dS m-1]
    initial_cumulative_transpiration = 0, # [mm]
    initial_cumulative_irrigation    = 0, # [mm]
    initial_cumulative_drainage      = 0  # [mm]
  )

  c(parms, drivers)
}

make_initial_state <- function(parms) {
  # names and ORDER must match the derivative vector returned by the model
  c(crop_dry_mass            = unname(parms$initial_crop_dry_mass),
    leaf_area_index          = unname(parms$initial_leaf_area_index),
    root_zone_water          = unname(parms$initial_root_zone_water),
    root_zone_salt           = unname(parms$initial_root_zone_salt),
    leaf_sodium              = unname(parms$initial_leaf_sodium),
    substrate_ph             = unname(parms$initial_substrate_ph),
    perceived_deficit        = unname(parms$initial_perceived_deficit),
    perceived_ec             = unname(parms$initial_perceived_ec),
    cumulative_transpiration = unname(parms$initial_cumulative_transpiration),
    cumulative_irrigation    = unname(parms$initial_cumulative_irrigation),
    cumulative_drainage      = unname(parms$initial_cumulative_drainage))
}
```

## Simulation driver (no plotting here)

```{r}
#| label: simulation

run_simulation <- function(parms, season_length = 150, time_step = 0.05) {
  # season_length [d]; time_step [d]  -- 0.05 d = 72 min, fine enough for the
  # fastest state here (the water sensor lag, 0.5 d)

  initial_state <- make_initial_state(parms = parms)
  time_sequence <- seq(from = 0, to = season_length, by = time_step)    # [d]

  solution <- deSolve::ode(y = initial_state,
                           times = time_sequence,
                           func = crop_water_salinity_model,
                           parms = parms,
                           method = "rk4")

  # deSolve already names the first column "time" [d]; keep it as is
  result_dt <- as.data.table(as.data.frame(solution))
  result_dt[]
}

run_scenarios <- function(season_length = 150) {
  # Sweeps the two formulation switches and the leaching policy, which is the
  # comparison the review argues actually matters.
  drivers <- make_drivers(season_length = season_length)

  scenario_grid <- CJ(stress_combination = c("multiplicative",
                                             "additive_potential",
                                             "non_equivalent"),
                      assimilation_mode = c("radiation", "transpiration"),
                      base_leaching_fraction = c(0.05, 0.20, 0.35),
                      sorted = FALSE)

  row_index <- 1  # <-- initialised for debugging
  scenario_list <- vector(mode = "list", length = nrow(scenario_grid))
  for (row_index in seq_len(nrow(scenario_grid))) {
    parms <- make_parms(
      drivers = drivers,
      stress_combination = scenario_grid[row_index, stress_combination],
      assimilation_mode  = scenario_grid[row_index, assimilation_mode],
      base_leaching_fraction = scenario_grid[row_index, base_leaching_fraction])

    one_run <- run_simulation(parms = parms, season_length = season_length)
    one_run[, stress_combination := scenario_grid[row_index, stress_combination]]
    one_run[, assimilation_mode  := scenario_grid[row_index, assimilation_mode]]
    one_run[, base_leaching_fraction := scenario_grid[row_index, base_leaching_fraction]]
    scenario_list[[row_index]] <- one_run
  }

  rbindlist(scenario_list)
}
```

## Visualisation (separate from the simulation)

```{r}
#| label: visualisation

plot_scenarios <- function(scenario_dt) {

  pList <- list()

  pList$p1 <- ggplot(scenario_dt,
                     aes(x = time, y = soil_ec,
                         colour = factor(base_leaching_fraction),
                         linetype = stress_combination)) +
    geom_line(linewidth = 0.7) +
    labs(x = "time [d]", y = "root zone EC [dS m-1]",
         colour = "leaching fraction [-]", linetype = "stress combination",
         title = "Salinity build-up under three leaching policies") +
    theme_bw()

  pList$p2 <- ggplot(scenario_dt,
                     aes(x = time, y = crop_dry_mass,
                         colour = stress_combination,
                         linetype = assimilation_mode)) +
    geom_line(linewidth = 0.7) +
    facet_wrap(~ base_leaching_fraction, labeller = label_both) +
    labs(x = "time [d]", y = "crop dry mass [g DM m-2]",
         title = "Formulation choice matters more than parameter values") +
    theme_bw()

  pList$p3 <- ggplot(scenario_dt,
                     aes(x = time, y = stress_stomatal, colour = stress_combination)) +
    geom_line(linewidth = 0.7) +
    geom_line(aes(y = stress_expansion), linetype = "dashed", linewidth = 0.5) +
    facet_wrap(~ base_leaching_fraction, labeller = label_both) +
    labs(x = "time [d]", y = "stress factor [-]",
         title = "Solid: stomatal stress. Dashed: expansion stress") +
    theme_bw()

  pList$p4 <- ggplot(scenario_dt,
                     aes(x = time, y = leaf_sodium_concentration,
                         colour = factor(base_leaching_fraction))) +
    geom_line(linewidth = 0.7) +
    labs(x = "time [d]", y = "leaf sap sodium [mmol_c L-1]",
         colour = "leaching fraction [-]",
         title = "Munns phase 2: the ionic integrator") +
    theme_bw()

  pList
}

run_all <- function(season_length = 150) {
  pList <- list()
  scenario_dt <- run_scenarios(season_length = season_length)
  pList <- plot_scenarios(scenario_dt = scenario_dt)
  list(scenario_dt = scenario_dt, pList = pList)
}
```

# Causal diagram

```{dot}
//| label: causal-diagram
digraph crop_water_salinity {
  rankdir = LR;
  graph [fontname = "Helvetica", nodesep = 0.35, ranksep = 0.55];
  node  [fontname = "Helvetica", fontsize = 10, style = filled];
  edge  [fontname = "Helvetica", fontsize = 9];

  subgraph cluster_drivers {
    label = "w(t) uncontrolled drivers"; color = grey70; style = dashed;
    node [shape = parallelogram, fillcolor = "#EEEEEE"];
    radiation      [label = "global radiation\n[MJ m-2 d-1]"];
    reference_et   [label = "reference ET\n[mm d-1]"];
    temperature    [label = "air temperature\n[degC]"];
    rainfall       [label = "rainfall\n[mm d-1]"];
    water_quality  [label = "irrigation water\nEC [dS m-1], alkalinity [mmol_c L-1]"];
  }

  subgraph cluster_states {
    label = "X states"; color = "#2166AC"; style = dashed;
    node [shape = box, fillcolor = "#D1E5F0"];
    dry_mass    [label = "crop dry mass\n[g DM m-2]"];
    leaf_area   [label = "leaf area index\n[m2 m-2]"];
    zone_water  [label = "root zone water\n[mm]"];
    zone_salt   [label = "root zone salt\n[g m-2]"];
    leaf_sodium [label = "leaf sodium\n[mmol_c m-2]"];
    substrate_ph [label = "substrate pH\n[-]"];
  }

  subgraph cluster_control {
    label = "u(t) = h_r(X, p_u, w)"; color = "#B2182B"; style = dashed;
    node [shape = box, fillcolor = "#FDDBC7"];
    perceived_deficit [label = "perceived deficit\n[-] sensor lag 0.5 d"];
    perceived_ec      [label = "perceived EC\n[dS m-1] lag 2 d"];
    irrigation        [label = "irrigation rate\n[mm d-1]"];
    leaching          [label = "leaching fraction\n[-]"];
    acid              [label = "acid dose\n[mmol_c m-2 d-1]"];
  }

  subgraph cluster_soil {
    label = "soil / substrate type p"; color = "#762A83"; style = dashed;
    node [shape = ellipse, fillcolor = "#E7D4E8"];
    retention  [label = "retention exponent\n[-] sand 4 .. clay 9"];
    buffering  [label = "pH buffer capacity\n[mmol_c m-2 pH-1]"];
  }

  node [shape = ellipse, fillcolor = "#D9F0D3"];
  matric   [label = "matric potential\n[MPa]"];
  osmotic  [label = "osmotic potential\n[MPa]"];
  stress_sto [label = "stomatal stress\n[-]"];
  stress_exp [label = "expansion stress\n[-]"];
  transpiration [label = "transpiration\n[mm d-1]"];
  assimilation  [label = "gross assimilation\n[g DM m-2 d-1]"];

  node [shape = doubleoctagon, fillcolor = "#FEE08B"];
  yield [label = "Y: yield [g DM m-2],\ndrain EC [dS m-1]"];

  radiation    -> assimilation;
  temperature  -> assimilation;
  reference_et -> transpiration;
  rainfall     -> zone_water;
  water_quality -> zone_salt;
  water_quality -> substrate_ph;

  zone_water -> matric;      retention -> matric;
  zone_salt  -> osmotic;     zone_water -> osmotic [label = "concentration"];
  matric  -> stress_sto;  osmotic -> stress_sto;
  matric  -> stress_exp;  osmotic -> stress_exp;

  stress_sto -> transpiration;
  stress_sto -> assimilation;
  stress_exp -> leaf_area [label = "fast, days"];
  leaf_area  -> assimilation [label = "1-exp(-k LAI)"];
  assimilation -> dry_mass;
  dry_mass -> assimilation [label = "maintenance, -", color = "#B2182B"];
  dry_mass -> leaf_area;

  transpiration -> zone_water [label = "-"];
  transpiration -> zone_salt  [label = "concentrates, +", color = "#B2182B"];
  transpiration -> leaf_sodium [label = "mass flow"];
  leaf_sodium -> leaf_area [label = "senescence, weeks", color = "#B2182B"];

  substrate_ph -> assimilation [label = "Fe/Mn, weeks"];
  buffering -> substrate_ph;

  zone_water -> perceived_deficit;  zone_salt -> perceived_ec;
  perceived_deficit -> irrigation;  perceived_ec -> leaching;
  leaching -> irrigation;
  irrigation -> zone_water;  irrigation -> zone_salt;
  leaching -> zone_salt [label = "flushes, -"];
  substrate_ph -> acid;  acid -> substrate_ph [label = "-"];

  dry_mass -> yield;  zone_salt -> yield;
}
```

View it from R:

```{r}
#| label: view-diagram

# DiagrammeR::grViz(readLines("crop_water_salinity.dot") |> paste(collapse = "\n"))
# or inline:
# DiagrammeR::grViz(dot_string)
```

The two red-labelled edges are the feedbacks that make this system interesting:
maintenance respiration (mass-proportional loss, the von Bertalanffy drag) and the
transpiration-to-salt-concentration edge, which is a **positive feedback under deficit
irrigation** — less water applied means less leaching means higher EC means more osmotic
stress. That loop is why deficit irrigation and saline water are a dangerous combination and
why the leaching fraction has to be a controlled variable rather than a fixed parameter.

# Gaps and recommendations

1. **Nobody carries the ionic integrator at field scale.** The Munns phase-2 state (leaf
   tissue sodium) is standard in physiology and absent from AquaCrop, WOFOST, DSSAT and
   STICS. Adding one state and two parameters would materially improve multi-season
   predictions under saline irrigation.
2. **Static salinity response functions are used dynamically.** Maas–Hoffman is a
   seasonal-yield regression; embedding it as an instantaneous factor double-counts, because
   the seasonal regression already includes the senescence effect that the dynamic model
   also computes. Letey & Feng make this point; it is still widely ignored.
3. **The combination rule is under-reported.** Multiplicative versus additive-potential
   versus non-equivalent changes predicted yield by a large margin under combined stress,
   and papers frequently do not state which they used. Always report it and always show
   sensitivity to it.
4. **pH is essentially unmodelled outside UNSATCHEM and EPIC.** For substrate systems this
   is a real loss, because pH there is fast, controllable and cheap to measure.
5. **Soil type enters in four independent places** (retention, salt storage, chemical
   buffering, sodicity-structure feedback) and most models represent only the first.
6. **Perception states are almost never modelled.** Irrigation-control papers optimise
   thresholds while assuming perfect, instantaneous measurement of soil water. Sensor lag
   and placement bias are first-order effects on the resulting policy.

# Glossary of technical abbreviations

Common ones (ODE, PDE, pH, EC, CO~2~) are not repeated here.

| Abbreviation | Expansion | Note |
|---|---|---|
| ABA | abscisic acid | root-to-shoot drought signal; the perception/transport lag is hours |
| AWC | available water capacity | mm of water per m of profile between field capacity and wilting point |
| CEC | cation exchange capacity | cmol~c~ kg^-1^; sets chemical buffering |
| DM | dry matter | biomass basis for all carbon states here |
| EC~e~ | EC of the saturated paste extract | the basis of Maas–Hoffman tables |
| EC~sw~ | EC of the soil solution at field water content | roughly 2x EC~e~; the quantity the plant sees |
| ESP | exchangeable sodium percentage | % of CEC occupied by Na^+^; sodicity measure |
| ET~0~ | reference evapotranspiration | FAO-56 grass reference, mm d^-1^ |
| FvCB | Farquhar–von Caemmerer–Berry | the C3 leaf photosynthesis biochemical model |
| GDD | growing degree days | degC d; thermal-time phenology driver |
| HI | harvest index | harvestable DM per total DM |
| K~cb~ | basal crop coefficient | transpiration-only crop coefficient, FAO-56 |
| LAI | leaf area index | m^2^ leaf per m^2^ ground |
| LF | leaching fraction | drainage volume / applied volume |
| LR | leaching requirement | the LF needed to hold root-zone EC below a target |
| LUE | light use efficiency | synonym of RUE in most crop-model usage |
| MOL | method of lines | spatial discretisation turning a PDE into an ODE system |
| PAR | photosynthetically active radiation | 400–700 nm, about 0.5 of global shortwave |
| PRD | partial root-zone drying | deficit-irrigation strategy alternating wetted side |
| RDI | regulated deficit irrigation | stage-targeted deficit strategy |
| RUE | radiation use efficiency | g DM per MJ intercepted PAR |
| SAR | sodium adsorption ratio | (mmol~c~ L^-1^)^0.5^; sodicity hazard of a solution |
| SLA | specific leaf area | m^2^ leaf per g leaf DM |
| TEC | threshold electrolyte concentration | EC below which clay disperses at a given SAR |
| VPD | vapour pressure deficit | kPa; the atmospheric side of the transpiration driver |
| WP* | normalised biomass water productivity | g DM m^-2^ mm^-1^, AquaCrop's central parameter |

Model names: **APEX** Agricultural Policy Environmental eXtender; **APSIM** Agricultural
Production Systems sIMulator; **CERES** Crop Environment REsource Synthesis; **CropSyst**
Cropping Systems Simulator; **DSSAT** Decision Support System for Agrotechnology Transfer;
**EPIC** Environmental Policy Integrated Climate; **GECROS** Genotype-by-Environment
interaction on CROp growth Simulator; **LINTUL** Light INTerception and UtiLization;
**SAHYSMOD** Spatial Agro-HYdro-Salinity MODel; **SALTMED** integrated irrigation, crop and
field-management model; **SIMPLE** the deliberately reduced crop model of Zhao et al.;
**STICS** Simulateur mulTIdisciplinaire pour les Cultures Standard; **SUCROS** Simple and
Universal CROp growth Simulator; **SWAP** Soil–Water–Atmosphere–Plant; **TOMGRO** tomato
growth model; **UNSATCHEM** unsaturated-zone chemistry model; **WOFOST** WOrld FOod STudies.

# References

Verify each before citing; recalled from training, not retrieved.

## Crop-model backbones

- van Ittersum, M.K. et al. (2003). On approaches and applications of the Wageningen crop
  models. *European Journal of Agronomy* 18: 201–234.
  <https://doi.org/10.1016/S1161-0301(02)00106-5>
- de Wit, A. et al. (2019). 25 years of the WOFOST cropping systems model.
  *Agricultural Systems* 168: 154–167. <https://doi.org/10.1016/j.agsy.2018.06.018>
- Spitters, C.J.T. & Schapendonk, A.H.C.M. (1990). Evaluation of breeding strategies for
  drought tolerance in potato by means of crop growth simulation. *Plant and Soil* 123:
  193–203. <https://doi.org/10.1007/BF00011268>
- Jones, J.W. et al. (2003). The DSSAT cropping system model. *European Journal of Agronomy*
  18: 235–265. <https://doi.org/10.1016/S1161-0301(02)00107-7>
- Keating, B.A. et al. (2003). An overview of APSIM. *European Journal of Agronomy* 18:
  267–288. <https://doi.org/10.1016/S1161-0301(02)00108-9>
- Holzworth, D.P. et al. (2014). APSIM – evolution towards a new generation of agricultural
  systems simulation. *Environmental Modelling & Software* 62: 327–350.
  <https://doi.org/10.1016/j.envsoft.2014.07.009>
- Brisson, N. et al. (2003). An overview of the crop model STICS. *European Journal of
  Agronomy* 18: 309–332. <https://doi.org/10.1016/S1161-0301(02)00110-7>
- Stockle, C.O. et al. (2003). CropSyst, a cropping systems simulation model. *European
  Journal of Agronomy* 18: 289–307. <https://doi.org/10.1016/S1161-0301(02)00109-0>
- Zhao, C. et al. (2019). A SIMPLE crop model. *European Journal of Agronomy* 104: 97–106.
  <https://doi.org/10.1016/j.eja.2019.01.009>
- Yin, X. & van Laar, H.H. (2005). *Crop Systems Dynamics: an ecophysiological simulation
  model for genotype-by-environment interactions (GECROS)*. Wageningen Academic.
- de Pury, D.G.G. & Farquhar, G.D. (1997). Simple scaling of photosynthesis from leaves to
  canopies without the errors of big-leaf models. *Plant, Cell & Environment* 20: 537–557.
  <https://doi.org/10.1111/j.1365-3040.1997.00094.x>

## Water-limited production

- Steduto, P. et al. (2009). AquaCrop – the FAO crop model to simulate yield response to
  water: I. Concepts and underlying principles. *Agronomy Journal* 101: 426–437.
  <https://doi.org/10.2134/agronj2008.0139s>
- Raes, D. et al. (2009). AquaCrop II: main algorithms and software description.
  *Agronomy Journal* 101: 438–447. <https://doi.org/10.2134/agronj2008.0140s>
- Hsiao, T.C. et al. (2009). AquaCrop III: parameterization and testing for maize.
  *Agronomy Journal* 101: 448–459. <https://doi.org/10.2134/agronj2008.0218s>
- Feddes, R.A., Kowalik, P.J. & Zaradny, H. (1978). *Simulation of Field Water Use and Crop
  Yield*. Pudoc, Wageningen. (Simulation Monographs.)
- Tardieu, F. & Davies, W.J. (1993). Integration of hydraulic and chemical signalling in the
  control of stomatal conductance and water status of droughted plants. *Plant, Cell &
  Environment* 16: 341–349. <https://doi.org/10.1111/j.1365-3040.1993.tb00880.x>
- Tuzet, A., Perrier, A. & Leuning, R. (2003). A coupled model of stomatal conductance,
  photosynthesis and transpiration. *Plant, Cell & Environment* 26: 1097–1116.
  <https://doi.org/10.1046/j.1365-3040.2003.01035.x>
- Steppe, K. et al. (2006). A mathematical model linking tree sap flow dynamics to daily stem
  diameter fluctuations and radial stem growth. *Tree Physiology* 26: 257–273.
  <https://doi.org/10.1093/treephys/26.3.257>
- Cowan, I.R. & Farquhar, G.D. (1977). Stomatal function in relation to leaf metabolism and
  environment. *Symposia of the Society for Experimental Biology* 31: 471–505.
- Medlyn, B.E. et al. (2011). Reconciling the optimal and empirical approaches to modelling
  stomatal conductance. *Global Change Biology* 17: 2134–2144.
  <https://doi.org/10.1111/j.1365-2486.2010.02375.x>
- Sperry, J.S. et al. (2017). Predicting stomatal responses to the environment from the
  optimization of photosynthetic gain and hydraulic cost. *Plant, Cell & Environment* 40:
  816–830. <https://doi.org/10.1111/pce.12852>
- Cochard, H. et al. (2021). SurEau: a mechanistic model of plant water relations under
  extreme drought. *Plant, Cell & Environment* / *Annals of Forest Science*.
  <https://doi.org/10.1111/pce.14192>

## Salinity, EC and combined stress

- Maas, E.V. & Hoffman, G.J. (1977). Crop salt tolerance – current assessment. *Journal of
  the Irrigation and Drainage Division ASCE* 103: 115–134.
- van Genuchten, M.Th. & Hoffman, G.J. (1984). Analysis of crop salt tolerance data. In:
  *Soil Salinity under Irrigation*, Springer.
- van Genuchten, M.Th. (1987). *A Numerical Model for Water and Solute Movement in and below
  the Root Zone*. USDA-ARS Research Report 121.
- Munns, R. (1993). Physiological processes limiting plant growth in saline soils: some
  dogmas and hypotheses. *Plant, Cell & Environment* 16: 15–24.
  <https://doi.org/10.1111/j.1365-3040.1993.tb00840.x>
- Munns, R. & Tester, M. (2008). Mechanisms of salinity tolerance. *Annual Review of Plant
  Biology* 59: 651–681. <https://doi.org/10.1146/annurev.arplant.59.032607.092911>
- Homaee, M., Feddes, R.A. & Dirksen, C. (2002). A macroscopic water extraction model for
  non-uniform transient salinity and water stress. *Soil Science Society of America Journal*
  66: 1764–1772. <https://doi.org/10.2136/sssaj2002.1764>
- Skaggs, T.H., van Genuchten, M.Th., Shouse, P.J. & Poss, J.A. (2006). Macroscopic
  approaches to root water uptake as a function of water and salinity stress. *Agricultural
  Water Management* 86: 140–149. <https://doi.org/10.1016/j.agwat.2006.06.005>
- Shani, U. & Dudley, L.M. (2001). Field studies of crop response to water and salt stress.
  *Soil Science Society of America Journal* 65: 1522–1528.
  <https://doi.org/10.2136/sssaj2001.6551522x>
- Shani, U., Ben-Gal, A., Tripler, E. & Dudley, L.M. (2007). Plant response to the soil
  environment: an analytical model integrating yield, water, soil type and salinity.
  *Water Resources Research* 43: W08418. <https://doi.org/10.1029/2006WR005313>
- Letey, J. & Feng, G.L. (2007). Dynamic versus steady-state approaches to evaluate
  irrigation management of saline waters. *Agricultural Water Management* 91: 1–10.
  <https://doi.org/10.1016/j.agwat.2007.02.014>
- Corwin, D.L., Rhoades, J.D. & Simunek, J. (2007). Leaching requirement for soil salinity
  control: steady-state versus transient models. *Agricultural Water Management* 90: 165–180.
  <https://doi.org/10.1016/j.agwat.2007.02.007>
- Ragab, R. (2002). A holistic generic integrated approach for irrigation, crop and field
  management: the SALTMED model. *Environmental Modelling & Software* 17: 345–361.
  <https://doi.org/10.1016/S1364-8152(01)00079-2>

## Soil physics, chemistry and structure

- van Genuchten, M.Th. (1980). A closed-form equation for predicting the hydraulic
  conductivity of unsaturated soils. *Soil Science Society of America Journal* 44: 892–898.
  <https://doi.org/10.2136/sssaj1980.03615995004400050002x>
- Simunek, J., van Genuchten, M.Th. & Sejna, M. (2016). Recent developments and applications
  of the HYDRUS computer software packages. *Vadose Zone Journal* 15(7).
  <https://doi.org/10.2136/vzj2016.04.0033>
- Suarez, D.L. & Simunek, J. (1997). UNSATCHEM: unsaturated water and solute transport model
  with equilibrium and kinetic chemistry. *Soil Science Society of America Journal* 61:
  1633–1646. <https://doi.org/10.2136/sssaj1997.03615995006100060014x>
- van Dam, J.C. et al. (2008). Advances of modeling water flow in variably saturated soils
  with SWAP. *Vadose Zone Journal* 7: 640–653. <https://doi.org/10.2136/vzj2007.0060>
- Quirk, J.P. & Schofield, R.K. (1955). The effect of electrolyte concentration on soil
  permeability. *Journal of Soil Science* 6: 163–178.
  <https://doi.org/10.1111/j.1365-2389.1955.tb00841.x>
- McNeal, B.L. (1968). Prediction of the effect of mixed-salt solutions on soil hydraulic
  conductivity. *Soil Science Society of America Proceedings* 32: 190–193.
- Ezlit, Y.D., Bennett, J.McL., Raine, S.R. & Smith, R.J. (2013). Modification of the McNeal
  clay swelling model. *Soil Science Society of America Journal* 77: 2149–2156.
  <https://doi.org/10.2136/sssaj2013.03.0097>
- Nye, P.H. (1981). Changes of pH across the rhizosphere induced by roots. *Plant and Soil*
  61: 7–26. <https://doi.org/10.1007/BF02277359>

## Greenhouse, soilless and control

- Sonneveld, C. & Voogt, W. (2009). *Plant Nutrition of Greenhouse Crops*. Springer.
  <https://doi.org/10.1007/978-90-481-2532-6>
- Carmassi, G., Incrocci, L., Maggini, R., Malorgio, F., Tognoni, F. & Pardossi, A. (2005).
  Modeling salinity build-up in recirculating nutrient solution culture. *Journal of Plant
  Nutrition* 28: 431–445. <https://doi.org/10.1081/PLN-200049163>
- Carmassi, G. et al. (2007). An aggregated model for water requirements of greenhouse tomato
  grown in closed rockwool culture with saline water. *Agricultural Water Management* 88:
  73–82. <https://doi.org/10.1016/j.agwat.2006.10.002>
- Massa, D., Incrocci, L., Maggini, R., Bibbiani, C., Carmassi, G., Malorgio, F. & Pardossi,
  A. (2011). Simulation of crop water and mineral relations in greenhouse soilless culture.
  *Environmental Modelling & Software* 26: 711–722.
  <https://doi.org/10.1016/j.envsoft.2011.01.004>
- Jones, J.W., Kenig, A. & Vallejos, C.E. (1999). Reduced state-variable tomato growth model.
  *Transactions of the ASAE* 42: 255–265. <https://doi.org/10.13031/2013.13214>
- Vanthoor, B.H.E., de Visser, P.H.B., Stanghellini, C. & van Henten, E.J. (2011). A methodology
  for model-based greenhouse design: part 2, description and validation of a tomato yield
  model. *Biosystems Engineering* 110: 378–395.
  <https://doi.org/10.1016/j.biosystemseng.2011.08.005>
- van Straten, G., van Willigenburg, G., van Henten, E. & van Ooteghem, R. (2010).
  *Optimal Control of Greenhouse Cultivation*. CRC Press.
  <https://doi.org/10.1201/b10321>

## Field-scale models with a pH pathway

- Williams, J.R. (1995). The EPIC model. In: Singh, V.P. (ed.) *Computer Models of Watershed
  Hydrology*, Water Resources Publications, 909–1000.
- Oosterbaan, R.J. SAHYSMOD: spatial agro-hydro-salinity model. ILRI, Wageningen.
  <https://www.waterlog.info/sahysmod.htm>
